Left-linear rewriting in adhesive categories

### 2024-04-23 concur/fusione.tex
minor

We can identify a large class of adhesive categories such that all
left-linear rewriting systems over such categories are strong
enforcing. Such class includes $\cat{Set}$ and it is closed
by comma and functor category constructions.  As such, it includes
essentially all categories (e.g., presheaves over set) that are
typically considered for modelling purposes. Notably, it contains the
category $\cat{Graph}$ of directed graphs.
This is a natural generalisation from adhesive to $\mathcal{M}$-adhesive catgeories of a class studied in \cite{baldan2011adhesivity} (see \Cref{app:fill} for details).

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-beahved class mentioned above. 
  .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (e.g., for graphs, nodes can be
merged while edges cannot). The idea here is to consider as graphs presheaves on categories in which there are object which plays the roles of \emph{roots}, i.e.~objects which are not codomains of any arrows besides the identity. \todo{D: quest'ultima frase ha senso?}%We first introduce signatures that are
%essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}[Root-preserving graphical rewriting systems]
	Let $\X$ be a category, an object $X\in \X$ is a \emph{root} if the only arrow with codomain $X$ is $\id{X}$. The category The
	category $\gph{X}$ of \emph{$\X$-graphs} is the category
	$\Set^{\X}$. A
	\emph{root-preserving graphical rewriting system} is a left-linear
	rewriting system $(\gph{X}, \R)$ such that  the
	following hold for each rule
	$(l\colon K\to L, r\colon K\to R)$ in $\R$
	\begin{enumerate}
		\item for every $X\in \X$ and $x\in L(X)$, there exists a root $Y$
		 and an arrow $f\colon Y\to X$ such that $x$ belongs to the image of
		$L(f)\colon L(Y)\to L(X)$;
		\item $r\colon K\to R$ is mono on the roots, i.e.~for every root $X\in \X$, the component $r_X:K(X)\to R(X)$ is injective.
	\end{enumerate}
\end{definition}

@@

Left-linear rewriting systems on $\cat{EGraphs}$, i.e.~in which the right half of a rule is mono but not regular in general, 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

>>>>

We can identify a large class of adhesive categories such that all
left-linear rewriting systems over such categories are strong
enforcing. Such class includes $\cat{Set}$ and it is closed
by comma and functor category constructions.  As such, it includes
essentially all categories (e.g., presheaves over set) that are
typically considered for modelling purposes. Notably, it contains the
category $\cat{Graph}$ of directed graphs.
This is a natural generalisation from adhesive to $\mathcal{M}$-adhesive categories of a class studied in \cite{baldan2011adhesivity} (see \Cref{app:fill} for details).

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-behaved class mentioned above. 
  .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

A first class consists of those rewriting systems over
possibly hierarchical graphical structures obtained as algebras of
suitable signatures where rules are constrained not to merge elements
of top level sorts in the hierarchy (e.g., for graphs, nodes can be
merged while edges cannot). The idea here is to consider graphical structures as 
presheaves on categories in which there are objects which play the role of 
\emph{roots}, i.e.~objects which are not codomains of any arrow besides 
the identity. \todo{D: quest'ultima frase ha senso? A: Ok. Ho cambiato ``as graphs'' con ``graphical structures''.}%We first introduce signatures that are
%essentially acyclic graphs whose nodes can be seen as sorts.

\begin{definition}[Root-preserving graphical rewriting systems]
	Let $\X$ be a category, an object $X\in \X$ is a \emph{root} if the only arrow with codomain $X$ is $\id{X}$. The
	category $\gph{X}$ of \emph{$\X$-graphs} is the category
	$\Set^{\X}$. A
	\emph{root-preserving graphical rewriting system} is a left-linear
	rewriting system $(\gph{X}, \R)$ such that  the
	following hold for each rule
	$(l\colon K\to L, r\colon K\to R)$ in $\R$
	\begin{enumerate}
		\item for every $X\in \X$ and $x\in L(X)$, there exists a root $Y$
		 and an arrow $f\colon Y\to X$ such that $x$ belongs to the image of
		$L(f)\colon L(Y)\to L(X)$;
		\item $r\colon K\to R$ is mono on the roots, i.e.~for every root $X\in \X$, the component $r_X:K(X)\to R(X)$ is injective.
	\end{enumerate}
\end{definition}

@@

Left-linear rewriting systems on $\cat{EGraphs}$,\todo{A: sostituirei ``, i.e. in which'' con ``where''. Giusto?} i.e.~in which the right half of a rule is mono but not regular in general, 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}
